Semi-supervised low-rank logistic regression for high-dimensional neuroimaging data

### 2015-05-31 paper_nips/paper.tex
finished first results section

\bigskip
\begin{figure}
\begin{centering}
\includegraphics[width=1.00\textwidth]{figures/accuracies.png}
\end{centering}
\caption{\textbf{Generalization performance}
Depicts in-sample (apply current model on training set, \textit{left column}),
out-of-sample (current model on test set, \textit{middle column}),
and out-of-dataset (compress other dataset using the
learned decomposition for classification, \textit{right column}) accuracy.
Curves for different choices of $\lambda$ as well as for ordinary
logistic regression (\textit{yellowish})
across iterations over the HCP task data (epochs).
XXX: Message ??
}
\end{figure}

\begin{figure}
\begin{centering}
\includegraphics[width=1.00\textwidth]{figures/figure_weights_perc75.png}
\end{centering}
\caption{\textbf{Interpretability of classification weights}
The voxel predictors corresponding to 5 exemplary
(of 18 total) psychological tasks (\textit{rows})
from the HCP dataset \cite{barch2013}.
\textit{Left column:} ordinary logistic regression (same
implementation but without bottleneck or autoencoder),
\textit{middle column:} semi-supervised low-rank logistic regression
($n=20$ latent components, $\lambda=0.5$, $\ell_1=0.1$, $\ell_2=0.1$),
\textit{right column:} voxel-wise average of whole-brain
activity maps from each task.
Low-rank logistic regression
a) puts higher absolute weights on relevant structure
b) lower ones on irrelevant structure,
and
c) yields BOLD-typical local contiguity.
All values are z-scored and thresholded at the 75th percentile.
}
\end{figure}

\begin{figure}
% XXX: figure not readable
\begin{centering}
\includegraphics[width=1.00\textwidth]{figures/semisup.png}
\end{centering}
\caption{\textbf{Relation between support recovery and out-of-dataset performance}
\textit{Upper row:} the fit-for-purpose of the classification models
is quantified by linear correlation between the
z-scored average task image and the z-scored weights from
ordinary (\textit{red}) and present (\textit{blue colors}) logistic
regression ($\lambda=0.25, 0.50, 0.75, 1.00$).
\textit{Lower row:} ordinary logistic regression is applied to the
ARCHI dataset after projection into the latent dimensions learned from
the HCP task data.
\textbf{whenever I tried to have 3+ times more rest than task images results dropped dramatically...naive interpretation: if the rest information is too rich it overrules task patterns and hinders the supervised learning
{\color{blue} Michael: Does this also happen when you double the task data?
i.e. rest:task 3:1, but duplicate the task data, yielding essentially 3:2?
I assume this will work, which again points to the instability of the network architecture}}
}
\end{figure}

@@

\begin{table}[h]
 \centering
 \resizebox{0.4\textwidth}{!}{%
 \begin{tabular}{lccc}
 \hline
 \textit{n}               & ICA + LogReg & SPCA + LogReg & SS-LR LogReg \\ \hline
 \multicolumn{1}{l|}{5}   & 37,53 \%     & 32,19 \%      & 96,50 \%     \\
 \multicolumn{1}{l|}{20}  & 80,98 \%     & 78,15 \%      & 97,33 \%     \\
 \multicolumn{1}{l|}{50}  & 84,19 \%     & 83,97 \%      & 97,69 \%     \\
 \multicolumn{1}{l|}{100} & 87,28 \%     & 82,19 \%      & 97,80 \%     \\ \hline
 \end{tabular}
 }
 \vspace{-0.2cm}
 \caption{Serial versus parallel dimensionality reduction and classification}
 \label{table_one}
 \end{table}

\paragraph{Model performance.}
\begin{itemize}
\item{20 components: high bias/low variance
100 compoentens: low bias/high variance}
\end{itemize}
\begin{itemize}
  \item{This was evaluated by the prediction
accuracy on a validation set (20\% of the training data) at each iteration
over the ARCHI task data}
  \item{the less lambda, the faster the convergence, with small differences
  though; smaller differences with larger number of components}
  \item{both 20 and 100 hidden components achieve 97\% out-of-sample
  accuracy}
  \item{ordinary logistic regression converges faster but with stagnation at
  92\% out-of-sample accuracy}
  \item{classification at chance 6\% in case of lambda==0 because 
  only the unsupervised layer of the estimator is optimized}
  \item{important: the thus only resting-state-trained decomposition achieves
  the same usefulness in the ARCHI classification than the
  decomposition derived in semi-supervised fashion}

out of sample performance: extract features of general relevance
beyond the supervised problem at hand
\end{itemize}

\begin{table}[h]
  \centering
  \resizebox{1.00\textwidth}{!}{%
\begin{tabular}{l|cccc|llllcccc|llll|}
  \hline
                            & \multicolumn{4}{c|}{\textit{n = 5}}                                                                                                                & \multicolumn{4}{c|}{\textit{n = 20}}                                                                                                                       & \multicolumn{4}{c}{\textit{n = 100}}                                                       \\ \hline
  \textit{}                 & \multicolumn{1}{l}{$\lambda=0.25$} & \multicolumn{1}{l}{$\lambda=0.50$} & \multicolumn{1}{l}{$\lambda=0.75$} & \multicolumn{1}{l|}{$\lambda=1.00$} & \multicolumn{1}{l}{$\lambda=0.25$}   & \multicolumn{1}{l}{$\lambda=0.50$}   & \multicolumn{1}{l}{$\lambda=0.75$}   & \multicolumn{1}{l|}{$\lambda=1.00$}   & $\lambda=0.25$   & $\lambda=0.50$   & $\lambda=0.75$   & \multicolumn{1}{l}{$\lambda=1.00$} \\ \hline
  Out-of-sample\\accuracy   & \textit{88,90\%}                   & 95,13\%                            & \textbf{96,49\%}                   & 95,72\%                             & 97,44\%                              & \textbf{97,80\%}                     & 97,33\%                              & \textit{97,32\%}                      & 97,21\%          & \textit{97,03\%} & \textbf{97,80\%} & 97,38\%                            \\
  Precision (mean)          & \textit{86,98\%}                   & 94,86\%                            & \textbf{96,28\%}                   & 95,36\%                             & \textbf{97,38\%}                     & 97,06\%                              & \textit{96,98\%}                     & 97,02\%                               & 96,90\%          & \textit{96,47\%} & \textbf{97,50\%} & 96,91\%                            \\
  Recall (mean)             & \textit{88,27\%}                   & 95,19\%                            & \textbf{96,57\%}                   & 95,70\%                             & \textbf{97,51\%}                     & 97,50\%                              & \textit{97,35\%}                     & 97,36\%                               & 97,20\%          & \textit{97,15\%} & \textbf{97,88\%} & 97,37\%                            \\
  F1 score (mean)           & \textit{86,64\%}                   & 94,89\%                            & \textbf{96,38\%}                   & 95,43\%                             & \textbf{97,42\%}                     & 97,22\%                              & \textit{97,11\%}                     & 97,13\%                               & 97,00\%          & \textit{96,71\%} & \textbf{97,65\%} & 97,07\%                            \\ \hline
  Out-of-dataset \\accuracy & 60,83\%                            & \textit{54,30\%}                   & 60,69\%                            & \textbf{62,92\%}                    & 79,72\%                              & \textbf{81,94\%} & 79,72\%           & \textit{79,44\%}                     & \textbf{82,08\%} & 81,66\%            & 81,25\%          & \textit{75,83\%}                   \\ \hline
  \end{tabular}

}
\end{table}

\paragraph{Feature identification.}
\begin{itemize}
  \item{fit for purpose; the right model should have
  high probability in the right places}
  \item{comparing to ordinary logistic regression, low-rank logreg yielded class weights 
  that were much more similar to feature of the respective training examples,
  for any choice of $\lambda$}
  \item{the higher $\lambda$, the better support recovery}
  \item{conversely, the higher lambda (i.e., bigger emphasis on the
  resting-state autoencoder), the more useful the low-dimensional transformations
  for classification in an independent dataset (ARCHI) as measured by f1 score
  for higher (i.e., 50 and 100) but not lower (i.e., 20) numbers
  of latent components}

support recovery:
Low-rank logistic regression
a) puts higher absolute weights on relevant structure
b) lower ones on irrelevant structure,
and
c) yields BOLD-typical local contiguity.

the features of the classes show that we captured activaiton properties,
not particpant or paradigm properties
\end{itemize}

>>>>

@@

\begin{table}[h]
 \centering
 \resizebox{0.5\textwidth}{!}{%
 \begin{tabular}{lccc}
 \hline
 \textit{n}               & ICA + LogReg & SPCA + LogReg & SS-LR LogReg \\ \hline
 \multicolumn{1}{l|}{5}   & 37,53 \%     & 32,19 \%      & 96,50 \%     \\
 \multicolumn{1}{l|}{20}  & 80,98 \%     & 78,15 \%      & 97,33 \%     \\
 \multicolumn{1}{l|}{50}  & 84,19 \%     & 83,97 \%      & 97,69 \%     \\
 \multicolumn{1}{l|}{100} & 87,28 \%     & 82,19 \%      & 97,80 \%     \\ \hline
 \end{tabular}
 }
 \vspace{-0.2cm}
 \caption{Serial versus parallel dimensionality reduction and classification}
 \label{table_one}
 \end{table}

\paragraph{Model performance.}
SS-LR LogReg was subsequently trained ($500$ epochs) across parameter choices
for the hidden components ($n=5, 20, 100$) and
the balance between autoencoder and logistic regression
($\lambda=0.25, 0.5, 0.75, 1$).
Assuming 5 latent directions of variation should yield models with
higher bias and smaller variance then SS-LR LogReg with 100 latent directions.
%
Given the 18-class problem of HCP, setting $\lambda$ to $0$
consistently yields generalization performance
at chance-level (5,6\%) because
only the unsupervised layer of the estimator is optimized.
%
At each epoch (i.e., iteration over the data),
the out-of-sample performance of the trained classifier
was assessed on 20\% of unseen HCP data.
Additionally, the ``out-of-dataset performance'' of the learned decomposition
was assessed by using it as dimensionality reduction of an
independent labaled dataset (i.e., ARCHI) and conducting ordinary
logistic regression on the ensuing component loadings.

\begin{table}[h]
  \centering
  \resizebox{1.00\textwidth}{!}{%
\begin{tabular}{l|cccc|cccc|cccc|cccc|}
  \hline
                            & \multicolumn{4}{c|}{\textit{n = 5}}                                                                                                            & \multicolumn{4}{c|}{\textit{n = 20}}                                                                                                                   & \multicolumn{4}{c}{\textit{n = 100}}                                                       \\ \hline
  \textit{}                 & \multicolumn{1}{l}{$\lambda=0.25$} & \multicolumn{1}{l}{$\lambda=0.5$} & \multicolumn{1}{l}{$\lambda=0.75$} & \multicolumn{1}{l|}{$\lambda=1$} & \multicolumn{1}{l}{$\lambda=0.25$}   & \multicolumn{1}{l}{$\lambda=0.5$}   & \multicolumn{1}{l}{$\lambda=0.75$}   & \multicolumn{1}{l|}{$\lambda=1$}   & $\lambda=0.25$   & $\lambda=0.5$    & $\lambda=0.75$   & \multicolumn{1}{l}{$\lambda=1$}    \\ \hline
  Out-of-sample\\accuracy   & \textit{88,90\%}                   & 95,13\%                            & \textbf{96,49\%}                  & 95,72\%                          & 97,44\%                              & \textbf{97,80\%}                    & 97,33\%                              & \textit{97,32\%}                   & 97,21\%          & \textit{97,03\%} & \textbf{97,80\%} & 97,38\%                            \\
  Precision (mean)          & \textit{86,98\%}                   & 94,86\%                            & \textbf{96,28\%}                  & 95,36\%                          & \textbf{97,38\%}                     & 97,06\%                             & \textit{96,98\%}                     & 97,02\%                            & 96,90\%          & \textit{96,47\%} & \textbf{97,50\%} & 96,91\%                            \\
  Recall (mean)             & \textit{88,27\%}                   & 95,19\%                            & \textbf{96,57\%}                  & 95,70\%                          & \textbf{97,51\%}                     & 97,50\%                             & \textit{97,35\%}                     & 97,36\%                            & 97,20\%          & \textit{97,15\%} & \textbf{97,88\%} & 97,37\%                            \\
  F1 score (mean)           & \textit{86,64\%}                   & 94,89\%                            & \textbf{96,38\%}                  & 95,43\%                          & \textbf{97,42\%}                     & 97,22\%                             & \textit{97,11\%}                     & 97,13\%                            & 97,00\%          & \textit{96,71\%} & \textbf{97,65\%} & 97,07\%                            \\ \hline
  Out-of-dataset \\accuracy & 60,83\%                            & \textit{54,30\%}                   & 60,69\%                           & \textbf{62,92\%}                 & 79,72\%                              & \textbf{81,94\%} & 79,72\%          & \textit{79,44\%}                     & \textbf{82,08\%} & 81,66\%         & 81,25\%          & \textit{75,83\%}                   \\ \hline
  \end{tabular}
}
\vspace{-0.2cm}
 \caption{Performance of SS-LR LogReg across model parameter choices}
  \label{table_two}
\end{table}

We made several noteworthy observations.
%
First, the purely supervised estimator ($\lambda=1$) achieved in no
instance the best accuracy, precision, recall or f1 scores on HCP data.
Classification by low-rank logistic regression is therefore facilitated by
imposing structure from the unlabeled rest data.
%
Second, the higher the number of latent components $n$,
the higher the out-of-dataset performance with small choices for $\lambda$.
This suggests that the presence of more rest-data-inspired hidden components
yields to more effective feature representation in unrelated task data.
%
Third, for $n=20$ and $100$ (but not $5$) the purely rest-data-trained
decomposition matrix ($\lambda=0$) resulted in
noninferior out-of-dataset performance
of X and Y, respectively (not shown in Table \ref{table_two}).
This confirms that guiding model learning by task-unrelated structure
extract features of general relevance
beyond the supervised problem at hand.

Depicts f1 scores for prediction of each of 38 classes across training epochs.
Ordinary logistic regression operating in voxel space (\textit{left plot})
converged faster but performed worse than
low-rank logistic regression for few
(\textit{middle plot}) and many (\textit{right plot}).
Autoencoder or rest data were not used for these analyses
($\lambda=1$).
Hence, projecting the input data into a reduced space for classification
yields higher class separability.

\paragraph{Individual effects of dimensionality reduction and rest data.}
We first quantified the gain of introducing a bottleneck layer
disregaring the autoencoder.
To this end, ordinary logstic regression was juxtaposed
with LR LogRed at $\lambda=1$ (no autoencoder).
For this experiment, we increased the difficulty of the classification problem
by including data from all 38 HCP tasks.
Indeed, increased class separability in component space,
as compared to voxel space, entails differences in f1 score of
$\approx{17\%}$ (Figure \ref{fig_dimred}).

\begin{figure}
\begin{centering}
\includegraphics[width=1.00\textwidth]{figures/accuracies.png}
\end{centering}
\vspace{-0.5cm}
\caption{\textbf{Effect of bottleneck in a 38-task classificaton problem}
Depicts f1 scores for prediction of each of 38 classes across training epochs.
Ordinary logistic regression operating in voxel space (\textit{left plot})
converged faster but performed worse than
low-rank logistic regression for few
(\textit{middle plot}) and many (\textit{right plot}).
Autoencoder or rest data were not used for these analyses
($\lambda=1$).
Hence, projecting the input data into a reduced space for classification
yields higher class separability.
}
\label{fig_dimred}
\end{figure}

We then quantified the impact of structure from rest data keeping
all model parameters constant
($n=20$, $\lambda=0.5$, $\ell_1=0.1$, $\ell_2=0.1$).
At the beginning of every epoch,
2000 task and 2000 rest maps were drawn with replacement
from same amounts of task maps but varying amounts of rest maps.
In frequently encountered data-scarce settings ($\approx{100}$ samples),
but not in data-rich settings ($\approx{1000}$ samples),
the repertoire of rest structure modulates model performance
(Figure \ref{fig_semisup}).

\begin{figure}
\begin{centering}
\includegraphics[width=1.00\textwidth]{figures/semisup_both.png}
\end{centering}
\vspace{-0.5cm}
\caption{\textbf{Effect of richness in rest data}
Gradient descent was performed on 2000 task and 2000 rest images.
These were drawn with replacement from identical versus varying
quantities of task versus rest maps, at the begining of each epoch.
}
\label{fig_semisup}
\end{figure}

\paragraph{Feature identification.}
We finally examined whether the models
were fit for purpose
(Figure \ref{fig_weights}).
%
To this end, we computed Pearson's correlation between the classifier weights
and the averaged neural activity map for each of the 18 tasks.
%
Ordinary logistic regression thus yielded a mean feature identification
of $\rho=0.28$.
%
For SS-LR LogReg ($\lambda=0.25, 0.5, 0.75, 1$),
a per-class-weight map was computed by matrix
multiplication of the two inner layers.
Feature identification thus ranged
between $\rho=0.35$ and $\rho=0.55$ for $n=5$,
between $\rho=0.59$ and $\rho=0.69$ for $n=20$, and
between $\rho=0.58$ and $\rho=0.69$ for $n=100$.
%
Consequently,
SS-LR LogReg puts higher absolute weights on relevant structure.
This reflect a higher signal-to-noise ratio, in part explained
by the more BOLD-typical local contiguity.
%
Conversely, SS-LR LogReg puts lower probability mass on irrelevant structure.
Despite lower interpretability, the
salt-and-pepper-like weight maps obtained from
ordinary logistic regression are sufficient for
good classification performance.
%
Hence, SS-LR LogReg yielded class weights 
that were much more similar to features of the respective training samples
for any choice of $\lambda$.
SS-LR LogReg captures genuine properties of neural activity patterns,
rather than participant- or paradigm-specific artefacts.

\begin{figure}
\begin{centering}
\includegraphics[width=0.98\textwidth]{figures/figure_weights_perc75.png}
\end{centering}
\vspace{-0.1cm}
\caption{\textbf{Classification weight maps}
The voxel predictors corresponding to 5 exemplary
(of 18 total) psychological tasks (\textit{rows})
from the HCP dataset \cite{barch2013}.
\textit{Left column:} ordinary logistic regression (same
implementation but without bottleneck or autoencoder),
\textit{middle column:} semi-supervised low-rank logistic regression
($n=20$ latent components, $\lambda=0.5$, $\ell_1=0.1$, $\ell_2=0.1$),
\textit{right column:} voxel-wise average of whole-brain
activity maps from each task.
Low-rank logistic regression
a) puts higher absolute weights on relevant structure
b) lower ones on irrelevant structure,
and
c) yields BOLD-typical local contiguity.
All values are z-scored and thresholded at the 75th percentile.
}
\label{fig_weights}
\end{figure}
